The complex structure and peculiar internal motions of the planetary nebula NGC 6210

### 2020-05-29 ms.tex
Tere changes from May

%%
%% Will macros
%%
% A better \ion command that works in more circumstances
\newcommand\ION[2]{#1\,\scalebox{0.9}[0.8]{\uppercase{#2}}}
\newcounter{ionstage}
\renewcommand{\ion}[2]{\setcounter{ionstage}{#2}% 
  \ensuremath{\mathrm{#1\,\scriptstyle\Roman{ionstage}}}}
\newcommand\hii{\ion{H}{2}}
\newcommand\nii{[\ion{N}{2}]}
\newcommand\oiii{[\ion{O}{3}]}
\newcommand\oii{[\ion{O}{2}]}
\newcommand\Wav[1]{\ensuremath{\lambda #1}}
% Chemical formulae
\newcommand*\chem[1]{\ensuremath{\mathrm{#1}}}

@@

\author[López et al.]{
  William J. Henney,\(^1\)\thanks{
    w.henney@irya.unam.mx,
    jal@astro.unam.mx,
    tere@astro.unam.mx,
    richer@astro.unam.mx
  }
  J. A. López\(^2\)\footnotemark[1]
  Ma.\ T. García-Díaz,\(^2\)\footnotemark[1]
  and M. G. Richer\(^2\)\footnotemark[1]
  \\
  \(^1\)\foreignlanguage{spanish}{
    Instituto de Radioastronomía y
    Astrofísica, Universidad Nacional Autónoma de México, Apartado
    Postal 3-72, 58090 Morelia, Michaoacán, Mexico}
  \\
  \(^2\)\foreignlanguage{spanish}{
    Instituto de Astronomía,
    Universidad Nacional Autónoma de México,
    Ensenada, Baja California, 22800, México}
}
% These dates will be filled out by the publisher
\date{Accepted XXX. Received YYY; in original form ZZZ}

@@

 \begin{figure}
\centering
  \includegraphics[width=0.4\textwidth]{tere-figs/Figure2a}
  \caption{Top panel shows the slit positions from the 2015 run, these are indicated as solid lines and 
labeled at the top against an [O III] HST image. Dashed lines refer to additional slit positions from other runs
and are labeled at the bottom of the image. The middle and bottom panels show the slit positions from 
the other runs placed at different position angles. }
  \label{fig:slit-positions}
\end{figure}

\begin{table}
\centering
\caption{OAN-SPM/Mezcal high-resolution spectroscopy of NGC~6210}
\label{table:pa5}
\begin{tabular}{l|cccccccc} \toprule
  &   Exp.~Time & Spectral  & Slit & P. A.   & slit        \\
  Epoch   &    (s) & Range  &  (\AA)    &    (\AA)  & (m$\mu$)     \\
  \midrule
1998/06/28          & 30 & \Ha  & 150 & 90 & O\\
1998/06/28          & 300 & \Ha  & 150 & 90 & N\\
1998/06/28           & 1200 & \Ha   & 150 & 90 & M,P,Q \\
2003/06/05         & 1800 &  \Ha   & 70 &0 &     A,B,D*,E*,F,H,I,K\\ 
2003/10/16   &  1800 &   \Ha    & 70 & $-$21 & T\\
2003/10/16    &  1800 &   \Ha   & 70 & $-$68 & V\\
2003/10/17    &  1800 &   \Ha  & 70 & 77 & W\\
2004/06/13   & 1800 &   \oiii  & 70 & $-$9 & R  \\  
2004/06/13   & 1800 &   \oiii  & 70 & $-$19 & S  \\  
2004/06/14    & 1800 &  \sii &150 & $-$19 & S  \\
2004/06/13   & 1800 &   \oiii  & 70 & $-$56 & U  \\  
2004/06/14   & 1800 &   \oiii  & 150 & 0 & A',L  \\
2011/05/21     & 1800 & \Ha  & 150 & 0 & G* \\
2011/05/21     & 600 & \oiii & 150 & 0  & G \\
2013/07/06    &  1800 & \Ha   & 150 & 0 & C,I,J \\
2015/08/18  &  1800 &   \Ha, \oiii  & 70 & 0 & C,D,E,F,G\\
2015/08/19  &  1800 &   \Ha, \oiii   & 70 & 0 & B,A*,I\\
2015/08/20   &  1800 &   \Ha, \oiii & 70 & 0 & H,J,K\\
2019/09/18  &  1800 &   \Ha    & 150 & 56 & \\
  \bottomrule
\end{tabular}
\end{table}

@@

\begin{figure*}
  \centering
  \includegraphics[width=1\textwidth]{tere-figs/Figure3}
  \caption{Position--velocity line profiles for the slit positions from the 2015 run are shown in sets where
the H$\alpha$ bi-dimensional line profile is on the left panel of the set, the \oiii{} line profile is on the central panel 
and the \nii{} line profile is on the right panel. The corresponding slit positions are indicated at the top of
each set of line profiles. The velocity scale is heliocentric and the spatial scale is indicated in arcsec 
with respect to the center of the nebula.}
  \label{fig:pv-array}
\end{figure*}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/turtle-halo-slit-x}
  \caption{
    Deep images and spectra of the outer halo of NGC~6210, showing knotty structure.
    (a)~Composite mosaic of \oiii{} images,
    each with an exposure time of 1800~s,
    giving a combined field of view of \(3' \times 3'\) centered on the nebula.
    The position of slit position X is shown, for which a deep \Ha{} spectrum was obtained.
    Note that different sections of the mosaic are shown with different brightness normalizations,
    leading to abrupt jumps at the boundaries. 
    (b)~\Ha{} spectrum from Slit~X, showing emission from two of the halo knots.
    The red dotted line shows the nebular systemic velocity.
  }
  \label{fig:halo-knots}
\end{figure}

\begin{figure}
\centering
%\includegraphics[width=0.7\textwidth]{tere-figs/Figure8}
%\includegraphics[width=0.327\textwidth]{tere-figs/opo9836f}
%\includegraphics[width=0.32\textwidth]{tere-figs/N6210_ruso}
\includegraphics[width=\linewidth]{sketch}
\caption{mages from the HST archive are shown on the bottom panels,  \oiii{} (left) and \nii{} (right), and their
corresponding images are shown in the top panels processed with a natural logarithm function and an edge 
detection algorithm. The most salient features in these images are indicated and labeled for reference.}
\label{fig:hst}
\end{figure}

@@

@@

\section{Proper motions}
\label{sec:proper-motions}

@@

The knot complexes are also visible in the \oiii{} spectra,
although they are fainter than the shells in this line.
There is also a slight difference in velocity,
with \oiii{} showing expansion velocities that are \num{2} to \SI{5}{km.s^{-1}} smaller than \nii{}.
If the expansion velocities within the knot complexes are assumed to increase with radius,
then this is evidence for ionization stratification along the line of sight
with the \oiii{} emission arising slightly closer to the star than \nii{}.
This is also consistent with the HST images (e.g., Figure~\ref{fig:hst}d),
which show that the \oiii{} emission associated with the complexes is more diffuse than the compact \nii{} knots.

@@

  \caption{Positions and velocities of nebular features in three dimensions}

@@

    l % Axis

@@

    NE Blue complex & 60 & \AxM{E} & -51 \pm 1 & 23 \pm 5 & 65 \pm 2 & -69 \pm 4 & 5 \pm 1 & 0.14 \pm 0.04 & 2.1 \pm 0.6\\

@@

    SW Red complex & 250 & \AxP{E} & +38 \pm 1 & 18 \pm 3 & 49 \pm 2 & 68 \pm 3 & 7 \pm 2 & 0.19 \pm 0.06 & 3.7 \pm 1.2\\
    N(E) Red complex & 30 & \AxP{E} & +25 \pm 2 & 5 \pm 4 & 30 \pm 2 & 81 \pm 7 & 4 \pm 1 & 0.26 \pm 0.21 & 8.0 \pm 6.7\\
    SW Faint Blue & 225 & \AxM{E} & -26 \pm 1 & 26 \pm 4 & 41 \pm 3 & -50 \pm 4 & 7 \pm 1 & 0.11 \pm 0.02 & 2.6 \pm 0.5\\

@@

    NW knot  & 310 & \AxP{E} & +18 \pm 1 & 11 \pm 8 & 24 \pm 4 & 63 \pm 17 & 8 \pm 1 & 0.18 \pm 0.10 & 6.7 \pm 4.5\\
    N(W) Red complex & 300 & \AxP{E} & +21 \pm 2 & 10 \pm 4 & 27 \pm 3 & 68 \pm 8 & 6 \pm 2 & 0.16 \pm 0.08 & 6.0 \pm 3.1\\

@@

    SE Blue complex & 135 &  \AxM{E} & -31 \pm 3 & 39 \pm 10 & 54 \pm 8 & -44 \pm 8 & 5 \pm 1 & 0.07 \pm 0.02 & 1.1 \pm 0.4\\
    SE knot & 135 & \AxM{E} & -35 \pm 3 & 39 \pm 10 & 57 \pm 7 & -47 \pm 8 & 8 \pm 1 & 0.09 \pm 0.02 & 1.5 \pm 0.4\\

@@

    (3)~Outflow axis that feature is best aligned with.

@@

For those nebular features where both proper motion and radial velocity measurements exist,
it is possible to estimate the full three-dimensional velocity vector.
Under the assumption that all motions are strictly radial,
one can also find the three-dimensional position.
These are given in Table~\ref{tab:3d} for a wide variety of nebular features
in the inner shells, intermediate shell, low-ionization knots, and outer lobes.\footnote{
  Note that the uncertainties given in the table are derived from observational errors only,
  and they do not include the systematic uncertainty in the distance.
}
When combining radial velocities and proper motion measurements,
it is necessary to account for the fact that radial velocities measure the material speed,
whereas proper motions measure pattern speeds of shocks or ionization fronts \citep{Mellema:2004a}. 
The exact correction factor (material speed divided by pattern speed)
is model-dependent (e.g., Appendix~A of \citealp{ODell:2009c}),
but is predicted to be generally larger than unity
\citetext{see Fig.~8 of \citealp{Jacob:2013a}}
unless the nebula is in the recombination phase.
For simplicity, we adopt a constant correction factor of 1.2 for all entries in Table~\ref{tab:3d},
which is consistent with the predicted value for expansion velocities greater than \SI{30}{km.s^{-1}}
\citetext{see Fig.~3 of \citealp{Schonberner:2019a}}.

@@

An approximate three-dimensional visualization of the nebular structure is given in Figure~\ref{fig:cut-axis-3d}.
The high-ionization shells have two distinct symmetry axes, A and B, which are close to orthogonal.
Axis~A is shared with the outer lobes but the lobes also present outflows along two additional axes, C and~D.
The low-ionization knot complexes and individual knots do not share any of these four axes,
with the notable exception of the N~knot, which lies along axis~A.

Despite the apparent complexity of the multi-axis structure, there are some clear systematic patterns,
one of which is illustrated in Figure~\ref{fig:inclinations}.
This shows histograms of the outflow inclination angle from the plane of the sky (absolute value)
for all of the individual features of Table~\ref{tab:3d}, classified by nebular component.
The inclination distribution for the outer lobes is very different from that of the knot complexes.
The lobes are concentrated at low inclinations, \(|i| = \ang{20 \pm 3}\),
whereas the knot complexes are concentrated at high inclinations, \(|i| = \ang{63 \pm 6}\).
The shells show intermediate inclinations, \(|i| = \ang{32 \pm 3}\),
while the individual knots are diverse
(the SE and NW~knot have high inclinations like the knot complexes,
while the N~knot has a low inclination like the lobes).

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/turtle-inclination-histogram}
  \caption{Distribution of outflow inclinations for the different components.}
  \label{fig:inclinations}
\end{figure}

The concentration of high inclinations for the knot complexes imply the presence of a fifth axis in the nebula, which we denote axis~E.
Since this axis is close to the plane of the sky,
its signature is not obvious on images of the nebula but can only be discerned from spectroscopy.
If we take the average vector outflow direction of the knot complexes,
then axis~E has an inclination \(i = \ang{69}\)
and a position angle \(\mathrm{PA} = \ang{331}\).
The individual knot complexes have outflows that deviate by only \ang{16} to \ang{29} from this average direction,
confirming that it is indeed a well-defined axis,
albeit with a substructure of three point symmetric pairs of outflows,
as discussed in \S~\ref{sec:knot-complexes}.

\begin{figure}
  \includegraphics[width=\linewidth]
  {figs/vel-radius-systems-annotated}
  \caption{
    True velocity, \(V_{\text{tot}}\), versus true radius, \(R\), for all the features listed in Table~\ref{tab:3d}.
    Diagonal lines show dynamical times between \num{1000} and \SI{16000}{years}, as marked.
    Symbol shape indicates the type of feature:
    circle symbols are high-ionization shells (\S~\ref{sec:high-ioniz-shells}),
    cross symbols are low-ionization knots or knot complexes (\S~\ref{sec:knot-complexes}),
    diamond symbols are high-velocity and intermediate-velocity systems in the outer lobes (\S~\ref{sec:outer-lobes}).
    Error bars show the observational uncertainties given in the table,
    except for outer lobe system \SysM{2}, where the lack of proper motion measurements mean that only lower limits for \(V_{\text{tot}}\) and \(R\) are available, indicated by arrows.
    The color of each symbol corresponds the position angle of the feature, as shown by the key at right,
    which also indicates the values for each of the four axes: A, B, C, and D.
    The position angle is calculated modulo \ang{180} so that the positive and negative arm of each axis have the same color.
    Also shown (gray ellipses) are the expansion velocity and range of radii of the inner and outer halo features (\S~\ref{sec:haloes}).
      }
  \label{fig:ages}
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]
  {figs/turtle-flow-axis-history}
  \caption{
    Ejection history of each flow axis, assuming no acceleration or deceleration of the features.
  }
  \label{fig:axis-history}
\end{figure}

\textbf{TODO Discussion of C and D being older than axis A}

@@

\begin{table*}
  \caption{Physical parameters of nebular components}
  \label{tab:summary}
  \begin{tabular}{
    l % Component
    r % Ha flux
    r % Density
    S % Mass
    r % ioniz param
    S % tstart
    S % tstop
    S % Mdot
    r % Vel
    }
    \toprule
    {}          & {\Ha{} flux} & {Ionized Density} & {Ionized Mass}    & {Ionization Parameter} & {\(t_{\text{start}}\)} & {\(t_{\text{end}}\)} & {\(\dot{M}\)} & {\(V\)}\\
    {Component} & {\% of total}& {\si{H.cm^{-3}}} & {\(M_\odot\)} &          & {\SI{1000}{yr}}      & {\SI{1000}{yr}}    & {\si{\msun.yr^{-1}}} & {\si{km.s^{-1}}}\\
    \midrule
    {(1)} & {(2)} & {(3)} & {(4)} & {(5)} & {(6)} & {(7)} & {(8)} & {(9)}\\
    \addlinespace
    Inner shells & 44\% & 4000--5000 & 0.078 & 0.0036--0.0070 & -4 & -1.5 & 3.1e-5 & 40--55\\
    Intermediate shell & 13\% & 700--1000 & 0.135 & 0.0080--0.0140 & -10 & -4 & 2.3e-5& 35--38\\
    Knot complexes & 42\% & 2000--5000 & 0.063 & 0.0004--0.0050 & -8 & -1.5 & 9.7e-6& 24--67\\
    Outer Lobes & 1\% & 160--320 & 0.024 & 0.0086--0.0224 & -5 & -2 & 8e-6 & 40--90\\
    Inner Halo & 0.5\% & 80--160 & 0.030 & 0.0200--0.0340 & -20 & -10 & 3e-6 & 15: \\
    Outer Halo & 0.2\% & 15--40 & 0.042 & 0.0540--0.0760 & -40 & -20 & 2.1e-6 & 7: \\
    \addlinespace
    NW knot & 0.01\% & 6700 & 1e-5 & 0.0006 & -7.1 & -6.7 & 2.5e-8 & 24\\
    SE knot & 0.08\% & 8200 & 6e-5 & 0.0034 & -1.25 & -0.95 & 2e-7 & 54\\
    N knot &  0.02\% & 3000 & 3e-5 & 0.0012 & -3.0 & -2.7 & 1e-7 & 67\\
    \bottomrule
  \end{tabular}
\end{table*}

@@

Once the density is known, then the column density of ionized hydrogen follows as proportional to \(S(\Ha)/n\) and 
the ionized mass of each component can be calculated as
\begin{equation}
  \label{eq:mass}
  M = \frac{4\pi D^2\, \bar{m}\, F(\Ha) }{\alpha_{\text{eff}}\, E(\Ha)\, n} \ , 
\end{equation}
where \(D\) is the distance from Earth (assumed \SI{2}{kpc}) and \(\bar{m} \approx \SI{2.17e-24}{g}\) is the mean mass per hydrogen nucleon.
The total ionized mass of the nebula is found to be \SI{0.372}{\msun} and results for individual components are shown in column~4 of Table~\ref{tab:summary}.
It is interesting to note that the high-density components (inner shell plus knot complexes)
only represent about one-third of the total ionized mass,
despite dominating the emission line flux.
The intermediate shell is the most massive single component (\SI{0.135}{\msun}),
also roughly one-third of the total,
with the remaining third corresponding to the outer lobes and halos.

\begin{figure}
  \includegraphics[width=\linewidth]
  {figs/mass-loss-history-annotated}
  \caption{
    Reconstruction of the mass-loss history of NGC~6210.
    Horizontal bars show the time period and mass-loss rate that led to the formation of each nebular component, as labeled.
    Only the average mass-loss rate is shown,
    although for the cases of the knot complexes and lobes the ejection was highly intermittent and non-isotropic during the indicated period.
    The red vertical line shows the end of the period of high mass-loss.
    Since then, the mass loss rate has declined to a much lower value. 
      }
  \label{fig:mass-loss-history}
\end{figure}

The masses can be combined with the kinematic ages (column~10 of Table~\ref{tab:3d}) to estimate the mass loss history of the nebula.
This is given in columns~6 to~8 of Table~\ref{tab:summary} and illustrated in Figure~\ref{fig:mass-loss-history}.
For the start and end times of each component, we take the full range of kinematic times of discrete features in the case of the lobes and knot complexes.
In the case of shells, the start time is taken as the kinematic age of material just outside the shell (for instance, the innermost halo material in the case of the intermediate shell).
In all cases, it is implicitly assumed that no significant acceleration or deceleration has taken place,
which may not be accurate, especially for the inner shells that have been shocked due to the action of the fast wind.
In general, the mass loss is seen to increase monotonically over the last \SI{40000}{yr} from about \SI{2e-6}{\msun.yr^{-1}} to about \SI{3e-5}{\msun.yr^{-1}} in the last few thousand years.
Note that the values shown in Figure~\ref{fig:mass-loss-history} are average values over the respective periods,
so the sudden jump that is apparent \SI{10,000}{yr} ago may have been a more gradual transition.
Also, this method can only estimate mass loss before \SI{1500}{yr} ago, which is the kinematic timescale of the inner shells.
Since that time, the mass-loss rate is expected to have declined significantly
due to the central star transitioning to higher effective temperature and smaller radius,
eventually reaching the present-day value of \(\dot{M} = \SI{9.12e-9}{M_\odot.yr^{-1}}\) at a wind velocity of \(V = \SI{2150}{km.s^{-1}}\) \citep{Herald:2011a}.
This is indicated on the figure by the gray downward pointing arrow.

@@

There is a marked difference in appearance of the nebula between
high-ionization lines such as \oiii{} (Fig.~\ref{fig:proper-motions-oiii})
and low-ionization lines such as \sii{}, \oii{} and \nii{} (Fig.~\ref{fig:proper-motions-nii}),
implying that there are strong variations in the degree of ionization.
The appearance in \Ha{} is broadly similar to \oiii{},
which suggests that the doubly ionized ions, such as \chem{O^{2+}} and \chem{N^{2+}},
are the dominant stages,
whereas the singly ionized ions, such as \chem{O^{+}} and \chem{N^{+}},
have low abundance in most of the nebula.
This in turn implies that the nebula must be optically thin in the \chem{H^0} and \chem{He^0} ionizing continua,
which is also consistent with the fact that the \nii{} emission is \emph{not} concentrated in an outer ring,
as would be expected for an optically thick nebula
\citetext{for example, the Ring Nebula, \citealp{ODell:2013b}}.

\newcommand\Fion{\ensuremath{F_{\text{ion}}}}
\newcommand\ionpar{\ensuremath{U_{\text{ion}}}}
In an optically thin nebula, the flux of ionizing photons at each point is \(\Fion = Q / 4\pi R^{2}\),
where \(Q\) is the ionizing photon luminosity.
The local ionization parameter is defined as \(\ionpar = \Fion / n c\), where \(c\) is the speed of light. 
We have calculated \(\ionpar\) for each of the features listed in Table~\ref{tab:3d}
and list in column~5 of Table~\ref{tab:summary} the range of values found for each component.
In this calculation, we use the density of each feature as calculated from the \Ha{} surface brightness (\S~\ref{sec:density-structure})
and assume \(Q = \SI{4.9e47}{s^{-1}}\), which we derive from the CSPN stellar atmosphere models of \citet{Krticka:2020a} using the stellar parameters of \citet{Herald:2011a}.
In Figure~\ref{fig:line-ratios} we plot the line ratios \nii{} \Wav{6583}/\Ha{} and \oiii{} \Wav{5007}/\Hb{} against the ionization parameter for each feature.
The \nii{} and \oiii{} surface brightness were determined from HST images, as outlined in \S~\ref{sec:density-structure}.

It is apparent from the table and figure that there is a wide range of ionization parameters in the nebula.
The steep decline in density with radius (faster than \(R^{-2}\)) means that the ionization parameter tends to increase with radius, although the lowest values are associated with the knot complexes, owing to their relatively high densities as compared with other features at similar radii.
Figure~\ref{fig:line-ratios} shows that the \nii/\Ha{} ratio varies by two orders of magnitude and is clearly anti-correlated with \ionpar.
The \oiii/\Hb{} ratio, on the other hand, shows no correlation with \ionpar{} and indeed varies little within the nebula.
This is what is expected in an optically thin nebula, where the doubly ionized metals predominate.

>>>>

%%
%% Will macros
%%
% A better \ion command that works in more circumstances
\newcommand\ION[2]{#1\,\scalebox{0.9}[0.8]{\uppercase{#2}}}
\newcounter{ionstage}
\renewcommand{\ion}[2]{\setcounter{ionstage}{#2}% 
  \ensuremath{\mathrm{#1\,\scriptstyle\Roman{ionstage}}}}
\newcommand\hii{\ion{H}{2}}
\newcommand\nii{[\ion{N}{2}]}
\newcommand\oiii{[\ion{O}{3}]}
\newcommand\Wav[1]{\ensuremath{\lambda #1}}

@@

\author[López et al.]{
  J. A. López,\(^1\)\thanks{
    jal@astro.unam.mx,
    tere@astro.unam.mx,
    w.henney@irya.unam.mx,
    richer@astro.unam.mx
  }
  Ma.\ T. García-Díaz,\(^1\)\footnotemark[1]
  William J. Henney\(^2\)\footnotemark[1]
  and M. G. Richer\(^1\)\footnotemark[1]
  \\
  \(^1\)\foreignlanguage{spanish}{
    Instituto de Astronomía,
    Universidad Nacional Autónoma de México,
    Ensenada, Baja California, 22800, México}
  \\
  \(^2\)\foreignlanguage{spanish}{
    Instituto de Radioastronomía y
    Astrofísica, Universidad Nacional Autónoma de México, Apartado
    Postal 3-72, 58090 Morelia, Michaoacán, Mexico}
}
% These dates will be filled out by the publisher
\date{Accepted XXX. Received YYY; in original form ZZZ}

@@

 \begin{figure}
\centering
  \includegraphics[width=0.4\textwidth]{tere-figs/Figure2a}
  \caption{ }
\end{figure}

\begin{table}
\centering
%\caption{Log  of Time-resolved  Observations of {\bf NGC~6210}.}
\begin{tabular}{l|cccccccc} \hline 
&  \multicolumn{7}{c}{Spectroscopy hight resolution/ 2.1m  OAN-SPM}            \\[0.1pt]
\hline 
Run    &   Exp. & Range  & Slit & P. A.   & slit        \\
Epoch   &     Time        &    (s)   &  (\AA)    &    (\AA)  & (m$\mu$)     \\[1pt]   \hline 
28/06/1998          & 30 & H$\alpha$  & 150 & 90 & O\\
28/06/1998          & 300 & H$\alpha$  & 150 & 90 & N\\
28/06/1998           & 1200 & H$\alpha$   & 150 & 90 & M,P,Q \\
05/06/2003         & 1800 &  H$\alpha$   & 70 &0 &     A,B,D*,E*,F,H,I,K\\ 
16/10/2003   &  1800 &   H$\alpha$    & 70 & $-$21 & T\\
16/10/2003    &  1800 &   H$\alpha$   & 70 & $-$68 & V\\
17/10/2003    &  1800 &   H$\alpha$  & 70 & 77 & W\\
13/06/2004   & 1800 &   O\sc{iii}  & 70 & $-$9 & R  \\  
13/06/2004   & 1800 &   O\sc{iii}  & 70 & $-$19 & S  \\  
13/06/2004   & 1800 &   O\sc{iii}  & 70 & $-$56 & U  \\  
14/06/2004   & 1800 &   O\sc{iii}  & 150 & 0 & A',L  \\
21/05/2011     & 1800 & H$\alpha$  & 150 & 0 & G* \\
21/05/2011     & 600 & O\sc{iii} & 150 & 0  & G \\
06/07/2013    &  1800 & H$\alpha$   & 150 & 0 & C,I,J \\
18/08/2015  &  1800 &   H$\alpha$, [OIII]  & 70 & 0 & C,D,E,F,G\\
19/08/2015  &  1800 &   H$\alpha$, [OIII]   & 70 & 0 & B,A*,I\\
20/08/2015   &  1800 &   H$\alpha$, [OIII] & 70 & 0 & H,J,K\\
18/09/2019  &  1800 &   H$\alpha$    & 150 & 56 & \\

\hline
\end{tabular}
\label{table:pa5}
\end{table}

@@

\begin{figure*}
\centering
  \includegraphics[width=1\textwidth]{tere-figs/Figure3}
  \caption{ }
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.4\textwidth]{tere-figs/Figure6}
  \caption{  }
\end{figure*}

\begin{figure}
\centering
%\includegraphics[width=0.7\textwidth]{tere-figs/Figure8}
%\includegraphics[width=0.327\textwidth]{tere-figs/opo9836f}
%\includegraphics[width=0.32\textwidth]{tere-figs/N6210_ruso}
\includegraphics[width=0.6\textwidth]{sketch}

  \caption{}
\end{figure}

@@

\clearpage

@@

\newpage
\section{Proper motions}
\label{sec:proper-motions}

@@

@@

  \caption{Components in three dimensions}

@@

    r % Axis

@@

    NE Blue complex & 60 &  & -51 \pm 1 & 23 \pm 5 & 65 \pm 2 & -69 \pm 4 & 5 \pm 1 & 0.14 \pm 0.04 & 2.1 \pm 0.6\\

@@

    SW Red complex & 250 &  & +38 \pm 1 & 18 \pm 3 & 49 \pm 2 & 68 \pm 3 & 7 \pm 2 & 0.19 \pm 0.06 & 3.7 \pm 1.2\\
    N(E) Red complex & 30 &  & +25 \pm 2 & 5 \pm 4 & 30 \pm 2 & 81 \pm 7 & 4 \pm 1 & 0.26 \pm 0.21 & 8.0 \pm 6.7\\
    SW Faint Blue & 225 &  & -26 \pm 1 & 26 \pm 4 & 41 \pm 3 & -50 \pm 4 & 7 \pm 1 & 0.11 \pm 0.02 & 2.6 \pm 0.5\\

@@

    NW knot  & 310 &  & +18 \pm 1 & 11 \pm 8 & 24 \pm 4 & 63 \pm 17 & 8 \pm 1 & 0.18 \pm 0.10 & 6.7 \pm 4.5\\
    N(W) Red complex & 300 &  & +21 \pm 2 & 10 \pm 4 & 27 \pm 3 & 68 \pm 8 & 6 \pm 2 & 0.16 \pm 0.08 & 6.0 \pm 3.1\\

@@

    SE Blue complex & 135 &  & -31 \pm 3 & 39 \pm 10 & 54 \pm 8 & -44 \pm 8 & 5 \pm 1 & 0.07 \pm 0.02 & 1.1 \pm 0.4\\

@@

    (3)~Outflow axis that feature is best aligned with, if any.

@@

For those nebular features where both proper motion and radial velocity measurements exist,
it is possible to estimate the full three-dimensional velocity vector.
Under the assumption that all motions are strictly radial,
one can also find the three-dimensional position.
These are given in Table~\ref{tab:3d} for a wide variety of nebular features
in the inner shells, intermediate shell, low-ionization knots, and outer lobes.

When combining radial velocities and proper motion measurements,
it is necessary to account for the fact that radial velocities measure the material speed,
whereas proper motions measure pattern speeds of shocks or ionization fronts \citep{Mellema:2004a}. 
The exact correction factor (material speed divided by pattern speed)
is model-dependent (e.g., Appendix~A of \citealp{ODell:2009c}),
but is predicted to be generally larger than unity
\citetext{see Fig.~8 of \citealp{Jacob:2013a}}
unless the nebula is in the recombination phase.
For simplicity, we adopt a constant correction factor of 1.2 for all entries in Table~\ref{tab:3d},
which is consistent with the predicted value for expansion velocities greater than \SI{30}{km.s^{-1}}
\citetext{see Fig.~3 of \citealp{Schonberner:2019a}}.

Note that the uncertainties given in the table are derived from observational errors only,
and they do not include the systematic uncertainty in the distance

@@

\begin{figure}
  \includegraphics[width=\linewidth]
  {figs/vel-radius-systems-annotated.pdf}
  \caption{
    True velocity, \(V_{\text{tot}}\), versus true radius, \(R\), for all the features listed in Table~\ref{tab:3d}.
    Diagonal lines show dynamical times between \num{1000} and \SI{16000}{years}, as marked.
    Symbol shape indicates the type of feature:
    circle symbols are high-ionization shells (\S~\ref{sec:high-ioniz-shells}),
    cross symbols are low-ionization knots or knot complexes (\S~\ref{sec:knot-complexes}),
    diamond symbols are high-velocity and intermediate-velocity systems in the outer lobes (\S~\ref{sec:outer-lobes}).
    Error bars show the observational uncertainties given in the table,
    except for outer lobe system \SysM{2}, where the lack of proper motion measurements mean that only lower limits for \(V_{\text{tot}}\) and \(R\) are available, indicated by arrows.
    The color of each symbol corresponds the position angle of the feature, as shown by the key at right,
    which also indicates the values for each of the four axes: A, B, C, and D.
    The position angle is calculated modulo \ang{180} so that the positive and negative arm of each axis have the same color.
    Also shown (gray ellipses) are the expansion velocity and range of radii of the inner and outer halo features (\S~\ref{sec:haloes}).
      }
  \label{fig:ages}
\end{figure}

@@

\begin{table*}
  \caption{Physical parameters of nebular components}
  \label{tab:summary}
  \begin{tabular}{
    l % Component
    r % Ha flux
    r % Density
    S % Mass
    r % ioniz param
    S % tstart
    S % tstop
    S % Mdot
    r % Vel
    }
    \toprule
    {}          & {\Ha{} flux} & {Ionized Density} & {Ionized Mass}    & {Ionization Parameter} & {\(t_{\text{start}}\)} & {\(t_{\text{end}}\)} & {\(\dot{M}\)} & {\(V\)}\\
    {Component} & {\% of total}& {\si{H.cm^{-3}}} & {\(M_\odot\)} & {\num{1e-4}}         & {\SI{1000}{yr}}      & {\SI{1000}{yr}}    & {\si{\msun.yr^{-1}}} & {\si{km.s^{-1}}}\\
    \midrule
    {(1)} & {(2)} & {(3)} & {(4)} & {(5)} & {(6)} & {(7)} & {(8)} & {(9)}\\
    \addlinespace
    Inner shells & 44\% & 4000--5000 & 0.078 & 36--70 & -4 & -1.5 & 3.1e-5 & 40--55\\
    Intermediate shell & 13\% & 700--1000 & 0.135 & 80--140 & -10 & -4 & 2.3e-5& 35--38\\
    Knot complexes & 42\% & 2000--5000 & 0.063 & 4--50 & -8 & -1.5 & 9.7e-6& 24--67\\
    Outer Lobes & 1\% & 160--320 & 0.024 & 86--224 & -5 & -2 & 8e-6 & 40--90\\
    Inner Halo & 0.5\% & 80--160 & 0.030 & 200--340 & -20 & -10 & 3e-6 & 15: \\
    Outer Halo & 0.2\% & 15--40 & 0.042 & 540--760 & -40 & -20 & 2.1e-6 & 7: \\
    \addlinespace
    NW knot & 0.01\% & 6700 & 1e-5 & 6 & -7.1 & -6.7 & 2.5e-8 & 24\\
    SE knot & 0.08\% & 8200 & 6e-5 & 34 & -1.25 & -0.95 & 2e-7 & 54\\
    N knot &  0.02\% & 3000 & 3e-5 & 12 & -3.0 & -2.7 & 1e-7 & 67\\
    \bottomrule
  \end{tabular}
\end{table*}

@@

In a similar way, the ionized mass of each component can be calculated as
\begin{equation}
  \label{eq:mass}
  M = \frac{4\pi D^2\, \bar{m}\, F(\Ha) }{\alpha_{\text{eff}}\, E(\Ha)\, n} \ , 
\end{equation}
where \(D\) is the distance from Earth (assumed \SI{2}{kpc}) and \(\bar{m} \approx \SI{2.17e-24}{g}\) is the mean mass per hydrogen nucleon.
The total ionized mass of the nebula is found to be \SI{0.372}{\msun} and results for individual components are shown in column~4 of Table~\ref{tab:summary}.
It is interesting to note that the high-density components (inner shell plus knot complexes)
only represent about one-third of the total ionized mass,
despite dominating the emission line flux.
The intermediate shell is the most massive single component (\SI{0.135}{\msun}),
also roughly one-third of the total,
with the remaining third corresponding to the outer lobes and halos.

\begin{figure}
  \includegraphics[width=\linewidth]
  {figs/mass-loss-history-annotated}
  \caption{
    Reconstruction of the mass-loss history of NGC~6210.
    Horizontal bars show the time period and mass-loss rate that led to the formation of each nebular component, as labeled.
    Only the average mass-loss rate is shown,
    although for the cases of the knot complexes and lobes the ejection was highly intermittent and non-isotropic during the indicated period.
      }
  \label{fig:mass-loss-history}
\end{figure}

The masses can be combined with the kinematic ages (column~10 of Table~\ref{tab:3d}) to estimate the mass loss history of the nebula.
This is given in columns~6 to~8 of Table~\ref{tab:summary} and illustrated in Figure~\ref{fig:mass-loss-history}.
For the start and end times of each component, we take the full range of kinematic times of discrete features in the case of the lobes and knot complexes.
In the case of shells, the start time is taken as the kinematic time of material just outside the shell (for instance, the innermost halo material in the case of the intermediate shell).
In all cases, it is implicitly assumed that no significant acceleration or deceleration has taken place,
which may not be accurate, especially for the inner shells that have been shocked due to the action of the fast wind.
In general, the mass loss is seen to increase monotonically over the last \SI{40,000}{yr} from about \SI{2e-6}{\msun.yr^{-1}} to about \SI{3e-5}{\msun.yr^{-1}} in the last few thousand years.
Note that the values shown in Figure~\ref{fig:mass-loss-history} are average values over the respective periods,
so the sudden jump that is apparent \SI{10,000}{yr} ago may have been a more gradual transition.
Also, the estimates are only for up to \SI{1500}{yr} ago, the kinematic timescale of the inner shells.
The mass-loss rate is expected to have declined significantly since then, with a present-day value of \(\dot{M} = \SI{9.12e-9}{M_\odot.yr^{-1}}\) at a wind velocity of \(V = \SI{2150}{km.s^{-1}}\) \citep{Herald:2011a}.

@@
